Toward single particle reconstruction without particle picking: Breaking the detection limit

### 2018-09-07 pnas/Towards.tex
no message

% Please add here a significance statement to explain the relevance of your work
\significancestatement{It is commonly believed that cryo-EM cannot be used to image the 3-D structure of small molecules (below ~50kDa). The classical argument is that, due to the high noise levels in cryo-EM, small particles cannot be detected reliably in micrographs. Challenging that belief, we demonstrate on a mathematical toy model that detection is not a prerequisite for structure reconstruction. Our reconstruction method is based on autocorrelation analysis. It requires only one pass over the data, so that it should, in principle, be appropriate for high-throughput experiments. This autocorrelation approach could also apply to other applications where the goal is estimation rather than detection.
}

@@

Figure~\ref{fig:Einst_example} shows several estimated images for a growing number of micrographs, and a movie is available as supplementary material. Figure~\ref{fig:error_per_micro} presents the normalized recovery error as a function of the amount of data available.  This is computed after fixing the reflection symmetries (see Section~\ref{sec:AC_analysis}). As evidenced by these figures, the ground truth image can be estimated increasingly well from increasingly many micrographs, without particle picking.

@@

\begin{figure}[h]
	\centering
	\includegraphics[scale=0.28]{Einstein_recovery_error_combined}
	\caption{\label{fig:error_per_micro}Relative error curves for the experiment of Figure~\ref{fig:Einst_example}.}
\end{figure}

\begin{figure*}[h!]
	\centering
	\includegraphics[scale=0.7]{1D_example}
	\caption{\label{fig:1D_example} Recovery of a 1-D signal of length $L=21$ at noise level $\sigma = 3$ from $M$ repetitions. The length of the micrograph was set to be $N = 410*M$. The first three panels (left to right) show reconstruction with different $M$ values compared to the ground truth signal (in blue).
		The true value of $\gamma$ is $0.0512$.
		   The relative errors of $\gamma$ for the three panels are: $4.8\%,4\%,1.2\%$. The right panel shows the relative error as a function of $M$. 
	}
\end{figure*}

@@

While we cannot provide a 3-D reconstruction from noisy data with the current algorithm, our method can be easily applied to the problem of deciding whether a micrograph contains projections or merely pure noise---a problem considered in classical works in statistics (e.g.,~\cite{donoho2004higher}) and debates in the cryo-EM community~\cite{henderson2013avoiding}.
This task can be performed by considering solely the recovered $\gamma$ (the fraction of pixels occupied by projections in the micrograph), estimated as part of the recovery algorithm.
Figure~\ref{fig:cryo_detection} presents excerpts  of two noisy micrographs, only one of which contains projections.
In the presence of projections, the estimated $\gamma$ was $0.12$, corresponding to approximately $6784$ projections. 
On the other hand, the estimated $\gamma$ drops to $10^{-5}$ for the pure noise micrograph, corresponding to less than one projection.

@@

@@

\paragraph{1-D experiment.}
For the 1-D experiment depicted in Figure~\ref{fig:1D_example}, we used a signal of length $L=21$ and generated an observation $y$ of length $N = 10(2L-1)M$, where $M$ is the number of repetitions. 
The observation was generated by randomly selecting placements in $y$, one at a time with an accept/reject rule based on the separation condition and locations picked so far until reaching the desired number of signal occurrences. Then, i.i.d.\ Gaussian noise with mean zero and standard deviation $\sigma = 3$ is added, to form the observed $y$. The resulting SNR of $y$ is about $1/175$.

@@

\paragraph{3-D experiments.}
\TODO{We need to add running times}
For the experiments in Figure~\ref{fig:cryo_detection}, we computed the analytical first three autocorrelations of the volume as explained in Appendix~\ref{sec:ac_cryo}.
To estimate the coefficients of the volume itself, we solve the LS problem
\begin{equation} \label{eq:LS_cryo}
\min_{\hat{x},\hat{\gamma}} w_1 \vert \mathfrak{a}_x^1 - \hat{\gamma}\mathfrak{a}_{\hat{x}}^1 \vert^2 + w_2\| \mathfrak{a}_x^2 - \hat{\gamma}\mathfrak{a}_{\hat{x}}^2 \|_2^2 + w_3\| \mathfrak{a}_x^3 - \hat{\gamma}\mathfrak{a}_{\hat{x}}^3 \|_F^2, 
\end{equation}  
where the explicit expressions of $\mathfrak{a}_x^1$, $\mathfrak{a}_x^2$ and $\mathfrak{a}_x^3$ are given in~\eqref{eq:ac_1_prolates},~\eqref{eq:ac_2_prolates} and~\eqref{eq:ac_3_prolates}, respectively.  
In the experiments, we set $w_1=w_2=w_3=1$. 
The LS problem was solved using Matlab's \texttt{lsqnonlin} solver for nonlinear LS problems. The expected $\hat \gamma$ in this experiment is one.

>>>>

% Please add here a significance statement to explain the relevance of your work
\significancestatement{It is commonly believed that cryo-EM cannot be used to image the 3-D structure of small molecules (below $\sim$50kDa). The classical argument is that, due to high noise levels in cryo-EM, small particles cannot be detected reliably in micrographs. Challenging that belief, we demonstrate on a mathematical toy model that detection is not a prerequisite for structure reconstruction. Our reconstruction method is based on autocorrelation analysis. It requires only one pass over the data, so that it should, in principle, be appropriate for high-throughput experiments. This autocorrelation approach could also apply to other applications where the goal is estimation rather than detection.
}

@@

Figure~\ref{fig:Einst_example} shows several estimated images for a growing number of micrographs, and a movie is available as supplementary material. Figure~\ref{fig:error_per_micro} (see appendix) presents the normalized recovery error as a function of the amount of data available.  This is computed after fixing the reflection symmetries (see Section~\ref{sec:AC_analysis}). As evidenced by these figures, the ground truth image can be estimated increasingly well from increasingly many micrographs, without particle picking.

@@

@@

While we cannot provide a 3-D reconstruction from noisy data with the current algorithm, our method can be easily applied to the problem of deciding whether a micrograph contains projections or merely pure noise---a problem considered in classical works in statistics (e.g.,~\cite{donoho2004higher}) and debates in the cryo-EM community~\cite{henderson2013avoiding}.
This task can be performed by considering solely the recovered $\gamma$ (the fraction of pixels occupied by projections in the micrograph), estimated as part of the recovery algorithm.
Figure~\ref{fig:cryo_detection} (see appendix) presents excerpts  of two noisy micrographs, only one of which contains projections.
In the presence of projections, the estimated $\gamma$ was $0.12$, corresponding to approximately $6784$ projections. 
On the other hand, the estimated $\gamma$ drops to $10^{-5}$ for the pure noise micrograph, corresponding to less than one projection.

@@

\begin{figure}[t]
	\centering
	\includegraphics[scale=0.28]{Einstein_recovery_error_combined}
	\caption{\label{fig:error_per_micro}Relative error curves for the experiment of Figure~\ref{fig:Einst_example}.}
\end{figure}

@@

\paragraph{1-D experiment.}

\begin{figure*}[t]
	\centering
	\includegraphics[width=\linewidth]{1D_example}
	\caption{\label{fig:1D_example} Recovery of a 1-D signal of length $L=21$ at noise level $\sigma = 3$ from $M$ repetitions. The length of the micrograph was set to be $N = 410*M$. The first three panels (left to right) show reconstruction with different $M$ values compared to the ground truth signal (in blue).
		The true value of $\gamma$ is $0.0512$.
		The relative errors of $\gamma$ for the three panels are: $4.8\%,4\%,1.2\%$. The right panel shows the relative error as a function of $M$. 
	}
\end{figure*}

For the 1-D experiment depicted in Figure~\ref{fig:1D_example}, we used a signal of length $L=21$ and generated an observation $y$ of length $N = 10(2L-1)M$, where $M$ is the number of repetitions. 
The observation was generated by randomly selecting placements in $y$, one at a time with an accept/reject rule based on the separation condition and locations picked so far until reaching the desired number of signal occurrences. Then, i.i.d.\ Gaussian noise with mean zero and standard deviation $\sigma = 3$ is added, to form the observed $y$. The resulting SNR of $y$ is about $1/175$.

@@

\paragraph{3-D experiments.}
\TODO{We need to add description of computer + running times}
For the experiments in Figure~\ref{fig:cryo_detection}, we computed the analytical first three autocorrelations of the volume as explained in Appendix~\ref{sec:ac_cryo}.
To estimate the coefficients of the volume itself, we solve the LS problem
\begin{equation} \label{eq:LS_cryo}
\min_{\hat{x},\hat{\gamma}} w_1 \vert \mathfrak{a}_x^1 - \hat{\gamma}\mathfrak{a}_{\hat{x}}^1 \vert^2 + w_2\| \mathfrak{a}_x^2 - \hat{\gamma}\mathfrak{a}_{\hat{x}}^2 \|_2^2 + w_3\| \mathfrak{a}_x^3 - \hat{\gamma}\mathfrak{a}_{\hat{x}}^3 \|_F^2, 
\end{equation}  
where the explicit expressions of $\mathfrak{a}_x^1$, $\mathfrak{a}_x^2$ and $\mathfrak{a}_x^3$ are given in~\eqref{eq:ac_1_prolates},~\eqref{eq:ac_2_prolates} and~\eqref{eq:ac_3_prolates}, respectively.  
In the experiments, we set $w_1=w_2=w_3=1$. 
The LS problem was solved using Matlab's \texttt{lsqnonlin} solver for nonlinear LS problems. The expected $\hat \gamma$ in this experiment is one.
